New and simplified manual controls for projection and slice tours, with application to exploring classification boundaries in high dimensions

### 2022-07-11 paper/paper.Rmd
captions for fig 6 and 7

The problem with new simple method (Algorithm 1) is that the precise values for $V_m$ are not followed because the orthonormalisation will change them. There are numerous ways that this can be enforced, and details are provided in the Appendix.

## Manual control for slice center

To better explore the space we combine the manual controls for the projection with manual controls for slicing. A slice is defined by a projection, a center point that is anchoring it in the high-dimensional space and the slice thickness $h$ [@slicetour]. In the case of a single orthogonal direction on the projection plane we can pick a sequence of center points in steps along that direction to move the slice and fully cover the data space. This no longer works in higher-dimensional spaces, and we can think of picking one direction and shifting the slice along the component orthogonal to the projection plane. In practice we define the center point explicitly and calculate the slice display as explained in @slicetour. The current implementation allows the user to individually enter the components of the vector defining the center point, and updates the display by jumping to the new slice.

@@

@@

We have seen that starting from $A_1$ using the manual controls to change the contribution of the second variable we could find a clear separation boundary indicating the relation between this variable (bill depth) and the Gentoo penguin species. Instead of rotating to a different projection, we might also change the view by moving the slice along one direction in the 4D space. Here we will continue our exploration of the dependence on `bd` and move the slice defined by $A_1$ to either large or small values of the bill depth ($\pm 1.5$ after centering and scaling). We will label these slices as $S_1^{\pm}$. Here we will also look at slices of the observed data points, using a thicker slice ($h=1.5$) to capture enough points in a given view.

We start by a comparison of the two models and the data distribution in $S_1^{+}$, thus the slice is localized towards high values of bill depth in Fig. \ref{slice1p}. We can see that all three slices (the two models and the data) contain almost no points from the third class (green, Gentoo), and that the decision boundary between the two models is very similar.

\begin{figure*}[ht]
\centerline{\includegraphics[width=0.32\textwidth]{figures/slice1_p_rf.png}
\includegraphics[width=0.32\textwidth]{figures/slice1_p_lda.png}
\includegraphics[width=0.32\textwidth]{figures/slice1_p_data.png}}
\caption{...}.
\label{slice1p}
\end{figure*}

A more interesting comparison is found for $S_1^{-}$, thus the slice localized towards low values of bill depth, shown in Fig. \ref{slice1m}. The RF model (left) predicts all three species within this slice, with an interesting boundary for the third class (green, Gentoo). On the other hand the LDA model (middle) predominantly predicts the third class within the slice, this appears to be enforced through the linear structure of the model. Looking finally at the thick slice through the data we see that there are primarily observations from this class within the slice we can conclude that the two models have filled in the "empty" space (where we do not have any training observations) in very different ways and according to what we might expect given the model structure.

\begin{figure*}[ht]
\centerline{\includegraphics[width=0.32\textwidth]{figures/slice1_m_rf.png}
\includegraphics[width=0.32\textwidth]{figures/slice1_m_lda.png}
\includegraphics[width=0.32\textwidth]{figures/slice1_m_data.png}}
\caption{...}.
\label{slice1m}
\end{figure*}

>>>>

The problem with new simple method is that it is not faithful to the precise values for $V_m$ because the orthonormalisation will change them. There are numerous ways that this can be enforced, and details are provided in the Appendix.

## Manual control for slices

To better explore the space we combine the manual controls for the projection with manual controls for slicing. A slice is defined by a projection, a center point that is anchoring it in the high-dimensional space and the slice thickness $h$ [@slicetour].

### Shifting the center

In the case of a single orthogonal direction on the projection plane we can pick a sequence of center points in steps along that direction to move the slice and fully cover the data space. This no longer works in higher-dimensional spaces, and we can think of picking one direction and shifting the slice along the component orthogonal to the projection plane. In practice we define the center point explicitly and calculate the slice display as explained in @slicetour. The current implementation allows the user to individually enter the components of the vector defining the center point, and updates the display by jumping to the new slice.

@@

### Changing the thickness

@@

We have seen that starting from $A_1$ using the manual controls to change the contribution of the second variable we could find a clear separation boundary indicating the relation between this variable (bill depth) and the Gentoo penguin species. Instead of rotating to a different projection, we might also change the view by moving the slice along one axis in the 4D space. Here we will continue our exploration of the dependence on `bd` and move the slice defined by $A_1$, $S_1^c$ to either large positive or negative values ($c_{bd} = \pm 1.5$ after centering and scaling). We will label these slices as $S_1^{\pm}$. Here we will also look at slices of the observed data points, using a thicker slice ($h=1.5$) to capture enough points in a given view.

We start by a comparison of the two models and the data distribution in $S_1^{+}$, thus the slice is localized towards high values of bill depth in Figure \ref{slice1p}. We can see that all three slices (the two models and the data) contain almost no points from the third class (green, Gentoo), and that the decision boundary between the two models is very similar.

\begin{figure*}[ht]
\centerline{\includegraphics[width=0.32\textwidth]{figures/slice1_p_rf.png}
\includegraphics[width=0.32\textwidth]{figures/slice1_p_lda.png}
\includegraphics[width=0.32\textwidth]{figures/slice1_p_data.png}}
\caption{Shifting the slice center in the positive direction of bill depth produces regions that have no Gentoo (green). The two models have similar boundaries except that the RF is more of a step function.}
\label{slice1p}
\end{figure*}

\begin{figure*}[ht]
\centerline{\includegraphics[width=0.32\textwidth]{figures/slice1_m_rf.png}
\includegraphics[width=0.32\textwidth]{figures/slice1_m_lda.png}
\includegraphics[width=0.32\textwidth]{figures/slice1_m_data.png}}
\caption{Shifting the slice center in the negative direction of bill depth produces regions that are mostly Gentoo (green). The two models have very different boundaries: the nonlinear potential of RF can be seen here where still some subspaces would predict to be Adelie and Chinstrap.}
\label{slice1m}
\end{figure*}

A more interesting comparison is found for $S_1^{-}$, thus the slice localized towards low values of bill depth, shown in Figure \ref{slice1m}. The RF model (left) predicts all three species within this slice, with an interesting boundary for the third class (green, Gentoo). On the other hand the LDA model (middle) predominantly predicts the third class within the slice, this appears to be enforced through the linear structure of the model. Looking finally at the thick slice through the data we see that there are primarily observations from this class within the slice we can conclude that the two models have filled in the "empty" space (where we do not have any training observations) in very different ways and according to what we might expect given the model structure.

### 2022-07-11 paper/paper.tex
captions for fig 6 and 7

The problem with new simple method (Algorithm 1) is that the precise
values for \(V_m\) are not followed because the orthonormalisation will
change them. There are numerous ways that this can be enforced, and
details are provided in the Appendix.

\hypertarget{manual-control-for-slice-center}{%
\subsection{Manual control for slice
center}\label{manual-control-for-slice-center}}

To better explore the space we combine the manual controls for the
projection with manual controls for slicing. A slice is defined by a
projection, a center point that is anchoring it in the high-dimensional
space and the slice thickness \(h\) \citep{slicetour}. In the case of a
single orthogonal direction on the projection plane we can pick a
sequence of center points in steps along that direction to move the
slice and fully cover the data space. This no longer works in
higher-dimensional spaces, and we can think of picking one direction and
shifting the slice along the component orthogonal to the projection
plane. In practice we define the center point explicitly and calculate
the slice display as explained in \citet{slicetour}. The current
implementation allows the user to individually enter the components of
the vector defining the center point, and updates the display by jumping
to the new slice.

@@

@@

We have seen that starting from \(A_1\) using the manual controls to
change the contribution of the second variable we could find a clear
separation boundary indicating the relation between this variable (bill
depth) and the Gentoo penguin species. Instead of rotating to a
different projection, we might also change the view by moving the slice
along one direction in the 4D space. Here we will continue our
exploration of the dependence on \texttt{bd} and move the slice defined
by \(A_1\) to either large or small values of the bill depth
(\(\pm 1.5\) after centering and scaling). We will label these slices as
\(S_1^{\pm}\). Here we will also look at slices of the observed data
points, using a thicker slice (\(h=1.5\)) to capture enough points in a
given view.

We start by a comparison of the two models and the data distribution in
\(S_1^{+}\), thus the slice is localized towards high values of bill
depth in Fig. \ref{slice1p}. We can see that all three slices (the two
models and the data) contain almost no points from the third class
(green, Gentoo), and that the decision boundary between the two models
is very similar.

\begin{figure*}[ht]
\centerline{\includegraphics[width=0.32\textwidth]{figures/slice1_p_rf.png}
\includegraphics[width=0.32\textwidth]{figures/slice1_p_lda.png}
\includegraphics[width=0.32\textwidth]{figures/slice1_p_data.png}}
\caption{...}.
\label{slice1p}
\end{figure*}

A more interesting comparison is found for \(S_1^{-}\), thus the slice
localized towards low values of bill depth, shown in Fig. \ref{slice1m}.
The RF model (left) predicts all three species within this slice, with
an interesting boundary for the third class (green, Gentoo). On the
other hand the LDA model (middle) predominantly predicts the third class
within the slice, this appears to be enforced through the linear
structure of the model. Looking finally at the thick slice through the
data we see that there are primarily observations from this class within
the slice we can conclude that the two models have filled in the
``empty'' space (where we do not have any training observations) in very
different ways and according to what we might expect given the model
structure.

\begin{figure*}[ht]
\centerline{\includegraphics[width=0.32\textwidth]{figures/slice1_m_rf.png}
\includegraphics[width=0.32\textwidth]{figures/slice1_m_lda.png}
\includegraphics[width=0.32\textwidth]{figures/slice1_m_data.png}}
\caption{...}.
\label{slice1m}
\end{figure*}

>>>>

The problem with new simple method is that it is not faithful to the
precise values for \(V_m\) because the orthonormalisation will change
them. There are numerous ways that this can be enforced, and details are
provided in the Appendix.

\hypertarget{manual-control-for-slices}{%
\subsection{Manual control for slices}\label{manual-control-for-slices}}

To better explore the space we combine the manual controls for the
projection with manual controls for slicing. A slice is defined by a
projection, a center point that is anchoring it in the high-dimensional
space and the slice thickness \(h\) \citep{slicetour}.

\hypertarget{shifting-the-center}{%
\subsubsection{Shifting the center}\label{shifting-the-center}}

In the case of a single orthogonal direction on the projection plane we
can pick a sequence of center points in steps along that direction to
move the slice and fully cover the data space. This no longer works in
higher-dimensional spaces, and we can think of picking one direction and
shifting the slice along the component orthogonal to the projection
plane. In practice we define the center point explicitly and calculate
the slice display as explained in \citet{slicetour}. The current
implementation allows the user to individually enter the components of
the vector defining the center point, and updates the display by jumping
to the new slice.

@@

\hypertarget{changing-the-thickness}{%
\subsubsection{Changing the thickness}\label{changing-the-thickness}}

@@

We have seen that starting from \(A_1\) using the manual controls to
change the contribution of the second variable we could find a clear
separation boundary indicating the relation between this variable (bill
depth) and the Gentoo penguin species. Instead of rotating to a
different projection, we might also change the view by moving the slice
along one axis in the 4D space. Here we will continue our exploration of
the dependence on \texttt{bd} and move the slice defined by \(A_1\),
\(S_1^c\) to either large positive or negative values
(\(c_{bd} = \pm 1.5\) after centering and scaling). We will label these
slices as \(S_1^{\pm}\). Here we will also look at slices of the
observed data points, using a thicker slice (\(h=1.5\)) to capture
enough points in a given view.

We start by a comparison of the two models and the data distribution in
\(S_1^{+}\), thus the slice is localized towards high values of bill
depth in Figure \ref{slice1p}. We can see that all three slices (the two
models and the data) contain almost no points from the third class
(green, Gentoo), and that the decision boundary between the two models
is very similar.

\begin{figure*}[ht]
\centerline{\includegraphics[width=0.32\textwidth]{figures/slice1_p_rf.png}
\includegraphics[width=0.32\textwidth]{figures/slice1_p_lda.png}
\includegraphics[width=0.32\textwidth]{figures/slice1_p_data.png}}
\caption{Shifting the slice center in the positive direction of bill depth produces regions that have no Gentoo (green). The two models have similar boundaries except that the RF is more of a step function.}
\label{slice1p}
\end{figure*}

\begin{figure*}[ht]
\centerline{\includegraphics[width=0.32\textwidth]{figures/slice1_m_rf.png}
\includegraphics[width=0.32\textwidth]{figures/slice1_m_lda.png}
\includegraphics[width=0.32\textwidth]{figures/slice1_m_data.png}}
\caption{Shifting the slice center in the negative direction of bill depth produces regions that are mostly Gentoo (green). The two models have very different boundaries: the nonlinear potential of RF can be seen here where still some subspaces would predict to be Adelie and Chinstrap.}
\label{slice1m}
\end{figure*}

A more interesting comparison is found for \(S_1^{-}\), thus the slice
localized towards low values of bill depth, shown in Figure
\ref{slice1m}. The RF model (left) predicts all three species within
this slice, with an interesting boundary for the third class (green,
Gentoo). On the other hand the LDA model (middle) predominantly predicts
the third class within the slice, this appears to be enforced through
the linear structure of the model. Looking finally at the thick slice
through the data we see that there are primarily observations from this
class within the slice we can conclude that the two models have filled
in the ``empty'' space (where we do not have any training observations)
in very different ways and according to what we might expect given the
model structure.
